---
title: "HEART DISEASE ANALYSIS"
author: "Aryan Khokhar"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## INTRODUCTION

This dataset is about heart disease. The dataset used for this analysis can be downloaded from <https://www.kaggle.com/datasets/johnsmith88/heart-disease-dataset>.

Before beginning, let me first lay the roadmap for this analysis. So, I'll perform the following tasks on this dataset:  
* Cleaning the data (if needed)  
* Transforming the data OR Data Manipulation  
* Analysis  
* Conclusion  

Starting with installing few packages and loading the dataset and also removing all the previously stored variables.

```{r Removing previously stored variables}
rm(list=ls()) #Removes all previously stored variables
```

```{r Installing tidyverse}
library(tidyverse)
```

```{r Loading dataset}
data <- read.csv("heart.csv")
view(data)
```

The dataset has 14 variables(columns) and 1025 observations(rows). Here we can use few other functions to get a brief summary about the data.

Here is the name of the columns and the explanation of each variable as described in Kaggle.

1. age: The age of a person  
2. sex: The person’s gender(1 = male, 0 = female)  
3. cp: The types of chest pain experienced (Value 1: typical angina, Value 2: atypical angina, Value 3: non-anginal pain, Value 4: asymptomatic)  
4. trestbps: Resting blood pressure (mm Hg on admission to the hospital)  
5. chol: Cholesterol measurement in mg/dl  
6. fbs: Fasting blood sugar (if > 120 mg/dl, 1 = true; 0 = false)  
7. restecg: Resting electrocardiographic measurement (0 = normal, 1 = having ST-T wave abnormality, 2 = showing probable or definite left ventricular hypertrophy by Estes’ criteria)  
8. thalach: Maximum heart rate achieved  
9. exang: Exercise induced angina (1 = yes; 0 = no)  
10. oldpeak: ST depression induced by exercise relative to rest (‘ST’ relates to positions on the ECG plot)  
11. slope: the slope of the peak exercise ST segment (Value 1: upsloping, Value 2: flat, Value 3: downsloping)  
12. ca: The number of major vessels (0–3)  
13. thal: A blood disorder called thalassemia (1 = normal; 2 = fixed defect; 3 = reversable defect)  
14. target: Heart disease (0 = no, 1 = yes)

```{r Summary of Dataset}
head(data) #Shows first 6 rows of the data
tail(data) #Shows last 6 rows of the data
glimpse(data)
summary(data) #Statistical overview of the data
```

Since, our data is cleaned, so I'll directly move onto the next step ,i.e, Data Transformation.

## DATA TRANSFORMATION

There are a lot of things that can be done with a dataset like this. Also, it can be analyzed in so many different ways. So many different plots and tables can be generated to explain in different ways.

To begin with, it will be helpful if we transform our data in another table and extract and focus on those variables only which are more important.

So, we'll make changes in "target,sex,fbs,exang,cp,restecg,slope,ca,thal" variables and leave the rest as it is.

```{r Transforming data}
data2 <- data %>% 
  mutate(sex = if_else(sex == 1, "MALE", "FEMALE"),
         fbs = if_else(fbs == 1, ">120", "<=120"),
         exang = if_else(exang == 1, "YES", "NO"),
         cp = if_else(cp == 0, "TYPICAL ANGINA",
                      if_else(cp == 1,"ATYPICAL ANGINA",
                       if_else(cp == 2, "NON-ANGINAL PAIN", "ASYMPTOMATIC"))),
         restecg = if_else(restecg == 0, "NORMAL",
                            if_else(restecg == 1, "ABNORMALITY", "PROBALBLE OR DEFINITE")),
         slope = as.factor(slope),
         ca = as.factor(ca),
         thal = as.factor(thal),
         target = if_else(target == 1, "YES", "NO")
         ) %>% 
  mutate_if(is.character,as.factor) %>% 
  dplyr::select(target,sex,fbs,exang,cp,restecg,slope,ca,thal,everything())

view(data2)
glimpse(data2)
```

The transformed data is ready. Let's dive into data visualisation.

## ANALYSIS

We will start with a simple bar graph depicting the presence and absence of heart disease in individuals. We'll count how many individuals are suffering from heart disease and how many are not. 

```{r Bar Plot for target (heart disease)}
ggplot(data2, aes(x=data2$target, fill=data2$target))+
  geom_bar()+
  xlab("Heart Disease")+
  ylab("Count")+
  ggtitle("Presence & Absence of Heart Disease")+
  scale_fill_discrete(name="Heart Disease", labels=c("Absence","Presence"))
```

We can also show this result in percentage of individuals having and not having heart disease.

```{r Percentage of heart disease}
round(prop.table(table(data2$target)),2) #To check the %age 
```

Now moving forward we want to see how the number of patients are distributed in different age groups and to check that first we'll form the 'age_grp' column using our original data. This is because here we want the sum of 'target' column and since, in transformed data 'target' column is factor variable that is why we are using the original data in which 'target' is an integer variable.

```{r Age Group Table}
data$age_grp <- cut(data$age, breaks=seq(25, 77, 4))

target_by_age <- data %>% 
                  group_by(age_grp) %>% 
                  summarise(heart_patients = sum(target))
target_by_age
```

Now let's visualise our above findings so that it is easier to interpret what we have obtained.

```{r Age Group Graph}
data %>% 
  group_by(age_grp) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x=age_grp, y=n), fill='pink')+
  ggtitle("Age Analysis")+
  xlab("Age Group")+
  ylab("Total Patients")
```

Further we can form a contingency table showing proportion of males and females suffering from heart disease or not.

```{r Proportion of heart patients among males and females}
round(prop.table(table(data2$sex, data2$target)), 2)
```

We can also compare the blood pressure in male and female patients, further comparing the results for different kinds of chest pain using a boxplot.

```{r Compare blood pressure with chest pain}
data2 %>% 
  ggplot(aes(x=sex, y=trestbps))+
  geom_boxplot()+
  xlab("Sex")+
  ylab("BP (mm Hg)")+
  facet_wrap(~cp)
```

Again let us make another boxplot and now compare the cholesterol level on male and female and grouped by chest pain.

```{r Compare cholesterol with chest pain}
data2 %>% 
  ggplot(aes(x=sex, y=chol))+
  geom_boxplot()+
  xlab("Sex")+
  ylab("Cholesterol (mg/dl)")+
  facet_wrap(~cp)
```

It will be great if we could see how the variables are correlated with each other. For that we must install and load the "corrplot" package as well as the "ggplot2" package.

```{r Installing correlation package}
#install.packages("corrplot")
library(corrplot)
library(ggplot2)
```

We'll start by finding out the correlation among last 5 columns of our dataset "data2" then finally showcasing it using corrplot() function.

```{r Correlation}
cor_heart <- round(cor(data2[,10:14]),2) 
cor_heart

corrplot(cor_heart, method = "square", type = "upper")
```

## CONCLUSION

* From our analysis, we can see that 51% of the individuals are suffering from heart disease. But the difference is not that much large and therefore, we can conclude that the chances of getting infected by a heart disease is almost equal to that of not having any heart disease.   
* The age group 53-61 has maximum number of heart patients.  
* The contingency table shows that the chances of a female for having a heart disease is 22% whereas for a male it is 29%. That is males are less immune to heart related problems.
* Males having 'Asymptotic' chest pain(CP) have less bloop pressure(BP) than females and same can be concluded for 'Atypical Angina' as well as 'Typical Angina' CP. But in case of 'Non-Anginal Pain' males tends to have slightly higher BP as compared to females.  
* However, males with 'Atypical Angina', 'Typical Angina' and 'Non-Anginal' CP have higher cholesterol level but for 'Asymptotic' CP both of them have same level of cholesterol level.  
* Lastly, from the correlation plot we can see that age of an individual and trestbps(resting blood pressure), chol(cholesterol level), oldpeak(ST depression induced by exercise relative to rest), are positively correlated, that is as the age of a person increases, these all values also increases.  
Whereas, the age and thalach(maximum heart rate achieved) are inversely correlated to each other.  
Also, thalach(maximum heart rate achieved) and oldpeak(ST depression induced by exercise relative to rest) are negatively correlated.